---
title: "Exploring AR(1) Relationships in R"
output: github_document
---

# Navigation

1.  [Earthquakes](#quakes)

    * [Data Exploration](#data-exploration)
    * [ACF](#acf)
    * [Relationship between x(t) and x(t-1)](#relationship-between-xt-and-xt-1)
    * [ACF of Residuals](#acf-of-residuals)

2.  [Cardiovascular Mortality Rate](#cmort)

    * [Data Exploration](#data-exploration-1)
    * [First Differences](#first-differences)
    * [ACF](#acf-1)
    * [Relationship between x(t) and x(t-1)](#relationship-between-xt-and-xt-1-1)
    * [ACF of Residuals](#acf-of-residuals-1)

# Quakes {#quakes}

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
```

This data is the annual number of earthquakes worldwide with a magnitude greater than 7.0 on the seismic scale.

```{r}
library(dplyr)
library(ggplot2)
theme_set(theme_bw())

# Read data and turn into data frame
x <- scan('https://online.stat.psu.edu/stat510/sites/stat510/files/L01/quakes.dat')
df <- data.frame(time = 1:length(x), num_quakes = x)

# Create lag(1) variable
df <- df %>%
  arrange(time) %>%
  mutate(lag1 = lag(num_quakes, 1))
```

## Data Exploration

First we plot the time series

```{r}
ggplot(aes(x = time, y = num_quakes), data = df) +
  geom_line() +
  xlab('Time') +
  ylab('Number of Earthquakes')
```

## ACF

Then we plot the ACF of the lags

```{r}
acf(x = df$num_quakes, xlim = c(1,20))
```

This shows that we have significant lags at 1, 2, 3, 4, and 5, with 1 being the largest.

-   Significant ACF indicates that data are not independent over time

## Relationship between x(t) and x(t-1)

Then we plot the number of earthquakes against its lag(1). Let's look at the linear relationship between $x_t$ and its first lag $x_{t-1}$

```{r}
# Linear model of x ~ lag(1)
ar1fit <- lm(num_quakes ~ lag1, data = df)
df$pred <- predict(ar1fit, newdata = df)

summary(ar1fit)
```

The model shows statistical significance with an $R^2$ value of 0.2972 - indicating that the lag(1) explains about 30% of the variance in a given day's number of earthquakes can be explained by the number of earthquakes on the previous day.

Let's look at this in a line graph

```{r}
ggplot(data = df) +
  geom_point(aes(x = lag1, y = num_quakes)) +
  geom_line(aes(x = lag1, y = pred))
```

## ACF of residuals

Next we check the autocorrelation function of the residuals to check if the model has properly captured the time dependence - good models should leave uncorrelated residuals.

Significant ACF means the model is missing structure; residuals should behave like random noise (white noise).

```{r}
acf(ar1fit$residuals, xlim = c(1,20))
```

There are no significant ACF values for the residuals - meaning no ARIMA terms, more lags, or transformations should be necessary.

# Cmort {#cmort}

This data concerns the weekly cardiovascular mortality rate in Los Angeles County.

```{r}
x <- scan('https://online.stat.psu.edu/stat510/sites/stat510/files/L01/cmort.dat')
df <- data.frame(mortality_rate = x, time = 1:length(x))
```

## Data exploration

First we plot the time series

```{r}
ggplot(aes(x = time, y = mortality_rate), data = df) +
  geom_line() +
  xlab('Time') +
  ylab('Mortality Rate')
```

There is a slight downward trend, so the series may not be stationary.

## First differences

To create a (possibly) stationary series, we’ll examine the first differences.

```{r}
df$mortdiff <- c(NA, diff(df$mortality_rate))
```

Let's see what this looks like

```{r}
ggplot(aes(x = time, y = mortdiff), data = df) +
  geom_line() +
  xlab('Time') +
  ylab('First Difference')
```

This looks a lot more stationary

## ACF

Now we look at the autocorrelation function of the first difference

```{r}
acf(df$mortdiff[!is.na(df$mortdiff)], xlim = c(1, 20))
```

This looks like the pattern of an AR(1) with a negative lag 1 autocorrelation.

## Relationship between x(t) and x(t-1)

```{r}
df <- df %>%
  arrange(time) %>%
  mutate(lag1 = lag(mortdiff, 1))

ar1mod <- lm(mortdiff ~ lag1, data = df)
summary(ar1mod)
```

This shows a negative linear relationship between $x_t$ and its first lag. The R-squared value of 0.2568 means that the difference between the previous mortality rate of a day and the day before that explains about 26% of the variance in the difference between mortality rate of the present day and the previous day.

-   More concisely: how much the previous day's change explains the current day's change

## ACF of Residuals

Now we check the autocorrelation function of the residuals

```{r}
acf(ar1mod$residuals, xlim = c(1, 20))
```

No significant ACF of residuals - meaning the model has successfully captured the time dependence in the data, and the residuals now behave like white noise (i.e., uncorrelated, random error)
